Proportional multi-state multiple-cohort life table model

### 2018-04-06 MSLT/MSLTdoc.Rmd
Figure 3 description

Table 1 describes data needs for the PMSLT, here we expand on the data needs and mechanisms (Figure 3) to use the PMSLT approach in ITHIMR (Figure 1). Figure 3 is used as a reference in the development of the code.

![Proportional multi-state life-table model. *Three sections are presented in Figure 3: **Data input sources**, **Inputs PMSLT** and **PMSLT mechanisms**.* ](structure\Figure3.png)

####Data for example

The code below is to load the data used in the example provided in this code for Australia (#4). *idata* includes data by one-year age groups and sex for: population numbers, mortality rates, total years lived with disability rates, incidence, case fatality, disability weights and costs for included diseases (ischemic heart disease, ischemic stroke, diabates, breast cancer (women) and colon cancer), cost for all other diseases (excludes those for physical inactivity) and trend for incidence and case fatality. Here we do not use cost and trends data. *edata* is for physical activity prevalence per categories (health risk factor), *irr* are relative risks and **ee** is energy expenditure per physical activity categories.

```{r, echo=TRUE}
#4
idata <- read.csv("data/idata.csv", stringsAsFactors = F)
edata <- read.csv("data/edata.csv", stringsAsFactors = F)
irr <- read.csv("data/irr.csv", stringsAsFactors = F)
ee <- read.csv("data/ee.csv", stringsAsFactors = F)
```

@@

```{r, echo=TRUE}
#5
p_age_cohort <- c(22, 27, 32, 37, 42, 47, 52, 57, 62, 67, 72, 77, 82, 87, 92, 97)
p_sex <- c("males", "females")
p_disease <- c("ihd", "istroke", "diabetes", "colon_cancer", "breast_cancer")
p_intervention_effect <- 100
```

>>>>

Table 1 describes data needs for the PMSLT, here we expand on the data needs and mechanisms (Figure 3) to use the PMSLT approach in ITHIMR (Figure 1).

(NEED TO PREPARE: LONDON, SAO PABLO, DELHI, BAY AREA). DATA AVAILABILITY FROM ARE IN 5 YEARS UP TO 95+. gbd DATA IS NOT AVAILABLE AT THE CITY LEVEL, HOWEVER, SOME LEVEL OF DISSAGREAGATION IS AVAILABLE. ENGLAND: EAST MIDLANDS, EAST OF ENGLAND, GRETER LONDON, ETC. wE COULD USE GREATER LONDON. FOR EDINBURGH WE HAVE DATA FOR SCOTLAND.  FOR SAO PABLO, DATA IS FOR BRAZIL. MEXICO IS NOT LISTED, BUT GBD HAS DATA FOR DISTRITO FEDERAL. FOR DELHI THE DATA IS FOR INDIA. FOR ACCRA THE DATA IS FOR GHANA. FOR LOS ANGELES DATA IS FOR UNITED STATES, BUT WE CAN HAVE THIS FROM NEIL?

![Proportional multi-state life-table model. *Three sections are presented in Figure 3: **Data input sources**, **Inputs PMSLT** and **PMSLT mechanisms**. The color coding from Data inputs sources to Inputs PMSLT link sources with inputs for the PMSLT. Solid arrows represent final inputs and dashed-arrows represent intermediate inputs that need further processing. Purple coding means a process and green coding represent change in mortality and disability prevalence rates to modify the life table parameters. Black color coding with white color font represent final model outcomes. For both, the life table and disease life table, two sets of each are simulated, one for the baseline and the other for the scenario. **Data inputs** are from four sources: Sythetic population, GBD data (from the Institute of Health Metrics and Evaluations (IHME)), ITHIM calculations and Injuries model. The synthetic population provides population numbers by age-groups and sex which are used in the life table (**PMSLT Mechanisms**) with each age cohort simulated until they die or reach 100 years of age. Population numbers are also used to derive road death rates and road injuries rates from numbers for the baseline and scenario generated by the Injuries model. The difference in road deaths rates modifies all cause-mortatliy in the scenario life table and the change in injuries rates modifies total YLDs rates (hence, healthy life years). Data from GBD IHME includes: total YLD rates per one, disease specific YLDs (numbers) and prevalence to derive disability weights for the included diseases, all cause mortatliy rates per one, population numbers and diseases parameters (incidence, mortality and prevalence). Dismod II is used to derive internally consitent estimates incidence and case fatality rates from GBD IHME data, which are inputs of the disease specific life tables. Case fatality derive from Dismod II is the same for the baseline and scenario (however, could be changed) and incidence is different. For the baseline life tables incidence is as derived from Dismod II and for the scanrio is modified by the PIF (1-PIF). A change in incidence modifies prevalence and mortality in later years. The difference between prevalence and mortality rates between the baseline and scanrio then feed into the scenario life table.* ](structure\Figure3.png)

(NEED TO PREPARE: LONDON, SAO PABLO, DELHI, ACCRA, LOS ANGELES, EDINBURGH)). DATA AVAILABILITY FROM ARE IN 5 YEARS UP TO 95+. gbd DATA IS NOT AVAILABLE AT THE CITY LEVEL, HOWEVER, SOME LEVEL OF DISSAGREAGATION IS AVAILABLE. ENGLAND: EAST MIDLANDS, EAST OF ENGLAND, GRETER LONDON, ETC. wE COULD USE GREATER LONDON. FOR EDINBURGH WE HAVE DATA FOR SCOTLAND.  FOR SAO PABLO, DATA IS FOR BRAZIL. MEXICO IS NOT LISTED, BUT GBD HAS DATA FOR DISTRITO FEDERAL. FOR DELHI THE DATA IS FOR INDIA. FOR ACCRA THE DATA IS FOR GHANA. FOR LOS ANGELES DATA IS FOR UNITED STATES, BUT WE CAN HAVE THIS FROM NEIL?

The code below is to load the data used will be updated to the study cases.

WRITE CODE HERE TO GET DATA.

@@

# ```{r, echo=TRUE}
# #5
# p_age_cohort <- c(22, 27, 32, 37, 42, 47, 52, 57, 62, 67, 72, 77, 82, 87, 92, 97)
# p_sex <- c("males", "females")
# p_disease <- c("ihd", "istroke", "diabetes", "colon_cancer", "breast_cancer")
# p_intervention_effect <- 100
# ```

####Data preparation

The following code applies to the data used as an exmaple here. #6 transforms upper cases to lower cases in the sex variable. #7 and #8 is used to rename mortality_rate to mx (used in life table language). #9 is used to replace male and female in the exposure data to males and females to facilitate combination of data sets.

# ```{r, echo=TRUE}
# #6
# idata$sex <- tolower(idata$sex)
# #7
# idata$mx <- idata$mortality_rate
# #8
# idata$mortality_rate <- NULL
# #9
# edata$sex[edata$sex=="male"] <- "males"
# edata$sex[edata$sex=="female"] <- "females"
# ```

# ```{r, echo=TRUE}
# ##### Create new variable for 5_year population (age cohorts). Depends on the age-cohorts of interest (5-yrs here)
# 
# idata$five_year_population <- NA
# start_index <- 3
# index <- 1
# end_index <- 4
# for (i in 1:20){
#   if (i == 1)
#     index <- 1
#   else 
#     index <- ((i - 1)  * 5) + 1
#   if (index == 96){
#     end_index <- 5
#   }
#   cat("population sum ", start_index, "start index ", index, " and end index ", index + end_index, "\n")
#   idata$five_year_population[start_index] <- sum(idata$population[index:(index + end_index)])
#   start_index <- start_index + 5
# }
# 
# end_index <- 4
# start_index <- 104
# 
# for (i in 1:20){
#   if (i == 1)
#     index <- 102
#   else
#     index <- (((i - 1)  * 5) + 1) + 101
#   if (index == 197){
#     end_index <- 5
#   }
#   cat("population sum ", start_index, "start index ", index, " and end index ", index + end_index, "\n")
#   idata$five_year_population[start_index] <- sum(idata$population[index:(index + end_index)])
#   start_index <- start_index + 5
# }
# 
# ```
